Venn diagrams may indicate erroneous statistical approaches leading to artifacts in transcriptomic analyses

### 2022-01-16 mk4/response_to_editor.md

Thanks for your responses to addressing the reviewer's concerns. I would
like to ask the authors to address the following concerns regarding the
paper. These are essential before accepting the paper.

The authors state that "Here we show that both, the use of Venn diagrams to
find genes which are thought to be specific for a certain comparison, as
well as gene set enrichment analysis applied to such subsets is a fallacy.
A statistically correct approach involves testing for interaction."

To start with, no one uses "Venn diagrams to find genes which are thought to be
specific for a certain comparison". It is just a visualization tool.

**We do agree ; we have stated that ourselves in the
Discussion. We have now reformulated the Abstract and Introduction.**

@@

The flaw
is not in the Venn diagram but in how one could interpret it. Also, applying
gene set enrichment analysis to such subsets could be done even without the use
of Venn diagrams.

**We fully agree; we have already addressed this issue in the
"Discussion". However...**

@@

The title and abstract are misleading and more journalistic
than scientific.  An analogous example to how the authors presented the paper
would be to condemn the use of a knife for murders.

**We understand that point. However, we do observe a substantial fraction of
papers using Venn diagrams misinterpreting the results. To continue on the
metaphor, the given analogy would only work if every fourth use of a knife
resulted in a homicide. This point has already been addressed in the
Discussion. Nonetheless we have now removed references to Venn diagrams in
the title and in the abstract.**

@@

This style is misleading
and not acceptable for a scientific paper. The title and abstract should be
completely reformatted to focus on the underlying issue.

**We have completely rewritten the title and the abstract and hope that
they are now acceptable.**

The authors state that "We have noticed a widespread fallacy related to gene
set enrichment analysis" however, no single example (or reference) has been
provided.

**We do not want to single out a paper or point fingers at specific
authors.  However, given that the sources of all our calculations are
available from github, we placed there the list of links to papers which –
in our view – have incorrectly interpreted Venn diagrams. For example, the
file `literature_survey_sci_immunol.md` includes six (out of 14 analysed)
links to papers in Science Immunology which incorrectly interpreted Venn
diagrams. The file `literature_survey_scirep.md` contains 73 links to
papers from Scientific Reports (out of 238 reviewed). We now mention that
explicitely in "Methods".**

@@

Authors state that "Of the 431 genes significant in G1, but not in G2, 199
(46%) are significant in the full data set; of the 278 genes significant in G2,
but not in G1, 99 (36%) are significant in the full data set. Given that G1 and
G2 were sampled from the total population, and since the FDR was set to 0.05,
we can assume that at least between a third and a half of the genes that
appeared to be specific" in the initial analysis were, in fact, false negatives
in one of the comparisons." The conclusion made is based on the assumption that
there are no false positives when analyzing the whole dataset. 
This is not
justifiable.

**This conclusion is based on the assumption that the fraction of FP among
significant genes is controlled by the Benjamini-Hochberg procedure, in
other words, that the false discovery rate (FDR) is not more than 5% (as
stated in the quoted paragraph).**

Also, splitting samples of a dataset into two groups results in
smaller sample sizes for each group.

**We agree, but the total number of samples in the analysis is still 80,
with 40 samples per group and 20 per group/treatment combination. This is
not an unusually small sample size for this type of transcriptomic
analysis (see below for more on the chosen sample size). In fact, in your
2019 paper on the influence of sample size on gene set enrichments that was
the largest group size you have used.
Also, the design of our in
silico experiment *requires* splitting the data set into two randomly selected groups,
because that is the whole idea of the paper: take a situation in which 
there should not be any significant enrichments (because there are no
real differences between the groups) and show that with an incorrect (but
widely spread) procedure you will get convincing evidence for differences
between groups.**

@@

This comes with two major shortcomings:
(1) the test for differential expression of genes are more sensitive to noise
and outliers, resulting in false positives

**We find the claim that the number of false positives is higher for small
sample sizes surprising. Statistical theory tells us that the FDR is
controlled by the BH false discovery rate controlling procedure.  If BH
(which is a widely accepted method for the analysis of DE data) is correct,
we should not expect more than 5% FP among significant results
independently of sample size.  See Benjamini, Yoav, and Yosef Hochberg.
"Controlling the false discovery rate: a practical and powerful approach to
multiple testing." Journal of the Royal statistical society: series B
(Methodological) 57.1 (1995): 289-300. We are not aware of any sources that
would claim otherwise.**

, and (2) the test for differential
expression of genes might not have the power to reject the null hypothesis (no
differential expression), resulting in introducing false negatives.

**We fully agree that FNs are the core of the problem! In fact, we have
chosen the FDR of 0.05 to minimize the number of false negatives (since a
more conservative threshold results in higher false negative rate).**

**However, we know from other studies (and basic statistics) that even in
most favorable circumstances the FNR will likely be above 20%; and that in
case of transcriptomic studies, the FNR may be much larger (please see the
relevant fragment already included in the paper). In other words, it does
not matter whether one looks at 80 samples or 193 samples: FNs are bound to
occur. That is the reason – we claim – for the fact that the gene set
enrichments found in this manner are related to the studied problem.
However, we have addressed this isssue by testing different sample sizes
(see below).**

@@

Therefore,
the design of the experiment is biased in favour of the "desired" conclusion.

**We have mimicked an RNA-Seq experiment using a sample size
well above the median used by others precisely to avoid this allegation. We
have now included a relevant reference (Baccarella et al.) which includes a
literature survey on RNA sample size in real world data. In there you will
see that half of the studies used sample sizes sizes of 6 or less per
group.**

**Nonetheless, we think that showing how the artifacts depend on sample size
is, in itself, interesting, not least because it shows that sample size has
a direct effect on the number of observed artifacts: the larger the sample
size, the more artifacts are observed. We have now included a section
comparing the results for 10, 20 (our example) and 40 samples per
group/treatment combination (thus, total sample sizes of 40, 80 and 160).
The reason for this is, of course, that the more DEGs you observe, the more
powerful the hypergeometric test is.**

**For each sample size, we have run 100 replicates of the group sampling,
replicated the whole procedure and gathered all test results. Results are
shown on the new Fig. 4, C and D.**

**However, we also think that the core of the problem is the incorrect
statistical procedure (selecting genes by comparing significance with lack
thereof) which will not become more correct by increasing the sample size.**

@@

XXX

The authors use a small log-fold change value of 1 to define the
differential expression. This leads to a high number of false positives.

**Statistical theory (e.g. the above reference to the original Benjamini &
Hochberg paper) tells us that the false discovery rate in differential
expression analysis is controlled by the BH procedure, not by the log fold
change threshold, except incidentally (beause log fold change is correlated
to the p-value, thus setting a more conservative log2 fold change results
in eliminating most of the larger p-values). Genes with large fold changes
may still not be significant and vice versa, true positives may have small
log2 fold changes. Therefore you cannot predict what will be the precise
effect of raising the LFC threshold on FDR except in very broad terms,
whereas setting the FDR threshold gives you clear information about the
expected proportion of false positives in your results.**

**This is analogous to using an effect size filter (say, Cohen's d) on
results of a t-test. The p-value from the t-test controls the type I error
rate, but a statistically significant difference does not necessarily
indicate a biologically meaningfull effect. For this, we turn to the effect
size (see for example Sullivan, Gail M., and Richard Feinn. "Using effect
size—or why the P value is not enough." Journal of graduate medical
education 4.3 (2012): 279-282.).**

**Thus, the threshold serves to eliminate genes that do show statistically
significant change (that is, they are true positives), but which are not
biologically relevant (because the effect size, for which log2 fold change
is a proxy, is too small). The proportion of false positives is 5% even with no
log2 FC treshold.**

**Increasing the threshold can, of course, result in not observing any
enrichments (see Fig. 4, A and B). Indeed, it does for some replicates.
This is not surprising – without a sufficient number of DEGs, no
enrichments can be calculated using a hypergeometric test. However, even
though the effect of increasing log2 fold change may sometimes eliminate the
symptoms, it does not cure the disease – the core problem of erroneous
analysis of interactions remains, and many replicates have artifactual
enrichments even at high log fold change thresholds.**

**Nonetheless, we have compared the effect of the log2
fold change threshold on the results (Fig. 4).**

### 2022-01-16 mk4/manuscript.rmd
major rewrite

---
title: "Venn diagrams may indicate erroneous statistical approaches leading to artifacts in transcriptomic analyses"
author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
date: "`r Sys.Date()`"
output:
  html_document
outputoff:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

@@

Venn diagrams are frequently used to illustrate the results of
differential expression analysis and further
downstream functional analysis. Here we show that both, the use
of Venn diagrams to find genes which are thought to be specific for a
certain comparison, as well as gene set enrichment analysis applied to such
subsets is a fallacy.  A statistically correct approach involves testing for interaction.
Moreover, since genes which show a significant change in one condition but
not in another are likely to be false negatives in the latter case, these
genes considered to be specific for one condition are likely characteristic
for both of the compared groups, and enriched in functions related to the
research hypothesis. Thus, the combination of incorrect statistical
analysis and gene set enrichment analysis may result in particularly
misleading artifacts.

@@

```{r setup}
padj_thr <- 0.05
lfc_thr  <- 1
n_stratum <- 20 # 2 strata per condition 
```

## Introduction: Venn diagrams are a showcase of erroneous interpretation of interactions

Venn diagrams (VDs) are commonly used to visualize high throughput data
such as transcriptomic profiles. Frequently, a Venn diagram serves the
purpose of comparing the results from two experiments or two comparisons.
For example, a VD may illustrate up- or down-regulated genes in two strains
of mice upon infection, showing which genes are regulated in both strains,
and which are regulated only in one of them.  The VD serves as a basis for
the statement that the regulation of certain genes is "specific" for one
strain or another. A gene is, in this context, considered "specific" or
"unique" for a given comparison when it shows a statistically significant
difference in that comparison only.  We argue that not only is this
inference incorrect, but it may also lead to misleading – although
appealing – artifacts when combined with downstream analyses.

Consider a gene for which the expression has been analysed in four groups:
two different mouse strains (wildtype, WT and knockout, KO) and two different experimental conditions
(naive versus infected). We find that the gene expression significantly
differs between infected and naive KO animals, but that there is no
significant difference in the WT strain. Such a gene may be incorrectly
considered as "specific" to KO, and will be accordingly entered in a Venn
diagram.

However, this is a common fallacy [@nieuwenhuis2011erroneous], since lack
of statistical significance is not the same as testing for lack of
difference. In other words, the fact that we failed to detect a significant
difference in the WT does not mean that the difference is absent and
itself significantly different from the difference in the KO. This "difference of
differences" is known is statistics as interaction [@blalock1965theory].
Here, it is the interaction between strain and treatment.  In fact, the
obtained p-values might be just over the assumed p-value threshold in one, but just
under it in the other strain (e.g. 0.009 vs 0.011), with the difference in
both strains being essentially the same. The correct analysis is to test
the significance of interaction using an ANOVA model. Notably, in 2011
Nieuwenhuis et al. found that in over 150 papers in top neuroscience
journals where the authors could make this particular statistical error,
half of them did.

Using VDs to show genes "specific" for a condition amounts to
counting the times a comparison for a gene was statistically significant in
one condition, but not significant in another.
More disturbingly, VDs are a visual illustration of a procedure
that exacerbates this problem by applying a downstream analysis to
supposedly specific sets of genes. Several cells in a Venn diagram contain
numbers corresponding to features (such as genes), for which there was a significant test result in
one, but not in another comparison. The sets of genes in a Venn diagram cell may
then be subsequently analysed to test whether they share a particular
characteristics.  For example, gene set enrichment analysis may be used to
interpret the biological function of genes that are considered "specific" to one
condition. It turns out that due to the
fallacious nature of this procedure, it is likely to produce results that
appear to make sense in the given biological context.

In this paper, we illustrate this rather simple statistical statement with
a real world data set, demonstrating how choosing this sort of approach
results in apparently sound gene set enrichment results which are, in fact,
artifacts. Next, we dissect the underlying mechanism of how these artifacts
are generated.  Finally, we discuss alternative approaches.

@@

First, we have performed differential gene expression analysis for each of
the groups G1 and G2 separately using standard bioinformatic tools. For each comparison, we defined
genes with significantly different gene expression 
(differentially expressed genes, DEGs) 
by using an false discovery rate (FDR) threshold of 
`r padj_thr`
and absolute log~2~ fold change (LFC) threshold of
`r lfc_thr`. There were
`r sum(res$g1$DEG)` 
DE genes in the G1 group, and
`r sum(res$g2$DEG)`
in the G2 group. In total,
`r sum(res.merged$DEG.g1 & res.merged$DEG.g2)` 
DEGs were common for G1 and G2, 
`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
DEGs were significant in G1 only ("specific" for G1), and 
`r sum(!res.merged$DEG.g1 & res.merged$DEG.g2)` 
were significant in G2 only ("specific" for G2; see Fig. 1, panel *A*).
A naive interpretation of these results implies that there is a substantial
difference between these two groups of individuals, as evidenced by a small
overlap in commonly regulated genes, while the majority of DEGs is
significant in one comparison only.

@@

```{r GSEA}
common <- res.merged %>% filter(DEG.g1 & DEG.g2) %>% pull(SYMBOL)
gsea_res <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset)
})
gsea_res_full <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset, qval = Inf)
})
gsea_res.merged <- merge(gsea_res$g1, gsea_res$g2, by=c("ID", "Title"), suffixes=c(".g1", ".g2"), all=T)
```

To understand which
pathways are upregulated in each of the two groups, we used a standard generation I
gene set enrichment analysis – a hypergeometric test – on the DEGs in each
group. Gene sets for the gene set enrichment analysis were taken from the Gene Ontology (GO) database.
Gene sets with more than `r max_mod_size` or 
fewer than `r min_mod_size` genes were removed.
For each group, we have
selected only genes which are DEGs in that group, but not the other,
mimicking a naive approach for finding "specifically" regulated pathways.
Here, a similar picture
emerged. Just
`r nrow(gsea_res$g1)` gene sets were significantly enriched in G1, and
`r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. Again,
both the Venn diagram (Fig. 1, B) and the results of
enrichments (Fig. 1, C and D) suggest that there is a
fundamental difference between the groups, and that the groups have little
in common in their response to the virus.

@@

@@

In the example above, the groups have been randomly sampled from a larger
data set only once. To investigate whether the sampling had an impact on
the artifacts, we have repeated the above procedure for 100 replicates,
each containing a different set of samples randomly assigned to the two
groups.  Furthermore, we tested whether the selection of the log~2~ fold
change threshold
or different sample size might have an impact on the extent of the arising
artifacts. To this end, we have tested `r length(lfc_thresholds)` different
threshold values for the log~2~ fold changes and three sample sizes: 20, 40 or 80 samples
per group (corresponding to 10, 20 or 40 samples per group/treatment
combination).

**Fig. 3.Influence of log~2~ fold change treshold (A, C) and the sample
size (B, D) on the number of genes significant in one, but not in the other
group (A, B) and the number of significant gene sets found in one group, but
not the other (C, D).**  Top row shows the influence of log~2~ fold change
threshold for sample size of 40 per group (20 per group/treatment
combination), bottom row shows the influence of sample size for a log~2~
fold change threshold set to `r lfc_thr`.

```{r montecarlo,cache=TRUE,fig.width=5.5,fig.height=5.5}
## note: given that the monte carlo simulation is quite time consuming, we
## have included the code for it in another file, `montecarlo.R`. If trying
## to replicate this manuscript, please execute that file first to generate
## the necessary replicates.

@@

@@

These FNs will occur
at a much higher rate within the sets of DEGs defined by the
non-overlaping areas of the VDs, that is DEGs considered to be "specific"
for one group or other in a naive approach.
To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
were drawn (Fig. 4), comparing the 
`r sum(covar$disease == "no") `
healthy controls to 
`r sum(covar$disease == "SC2") `
COVID-19 patients. Of the
`r length(g1_spec)`
genes significant in G1, but not in G2, 
`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g1_spec) %>% nrow; .n`
(`r sprintf("%.0f%%", .n/length(g1_spec) * 100)`) 
are significant in the full data set; 
of the
`r length(g2_spec)`
genes significant in G2, but not in G1, 
`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g2_spec) %>% nrow; .n`
(`r sprintf("%.0f%%", .n/length(g2_spec) * 100)`) 
are significant in the full data set.
Given that G1 and G2 were sampled from the total population, and since the
FDR was set to `r padj_thr`, we can assume that at least between a third and a half of
the genes that appeared to be "specific" in the initial analysis were, in
fact, false negatives in one of the comparisons.

**Fig. 4. Area-proportional Venn diagram showing overlaps in DEGs between
G1, G2 and the full data set.** The majority of genes which have been labeled
as DEGs in only one of the groups G1, G2 are DEGs when all data were
analysed.

@@

In an informal survey of articles from three journals (Tab. 2) we found
that of the 
`r tab2$Analysed[nrow(tab2)]` analysed articles
 which used the term "venn diagram" and
"differential expression", at least 
`r tab2$Incorrect[nrow(tab2)]`
were using Venn diagrams to compare
statistical significance with lack thereof by referring to "unique",
"specific", "solely regulated" or "exclusive" DEGs. Out of these, at least 
`r tab$Enrichment[nrow(tab2)]`
coupled the VDs with some form of gene set enrichment analysis on the
set of supposedly "specific" DEGs.

Drawing conclusions from comparing significance with lack thereof is a
common statistical fallacy [@gelman2006difference]. Just as absence of evidence is not evidence of
absence, the failure to reject the null hypothesis does not consitute the
same level of evidence as rejecting it. However, when such an incorrect
analysis is combined with downstream functional analysis – i.e., gene set
enrichments of genes "specific" to one or another comparison – the
resulting pathways or gene ontologies are misleadingly relevant, yielding
gene sets associated with immune answer for research hypotheses involving
an infectious disease, or cancer pathways if the underlying research
hypothesis involved cancer treatment. Such results may be hard to resist
for an involved researcher, especially if the correct analysis of
interactions does not show any significant differences.

That is not to say that VDs are not a useful tool, even in the context of
transcriptomics and gene set enrichments, if used correctly. For example, analysis of an
intersection of DEGs (i.e., by considering genes from the overlap in a VD)
is not an incorrect procedure.  Genes in the overlapping part of a VD are
significant in both (or all) comparisons, hence no comparison between
significance and non-significance is made.

As an alternative to Venn diagrams and the downstream gene set enrichment
analysis, two approaches can be considered.

Statistical analysis, as shown above, involves a test for interaction which
can reveal genes for which the impact of treatment significantly differ
between the groups. The results can then be plugged into a gene set
enrichment analysis the usual way.  Unfortunately, this has two major
drawbacks. Firstly, effect sizes (log~2~ fold changes) of the interaction
term are harder to interpret than log~2~ fold changes in a direct, group vs
group comparison.  The effect size in an interaction is negative if the
log~2~ fold change in the first comparison is larger than the log~2~ fold
change in the second comparison; this is, however, irrespective of whether
the differences in the individual comparisons are negative or positive,
which makes it harder to separate the differences in genes up-regulated in
one or both groups.

>>>>

---
title: ""
author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
date: "`r Sys.Date()`"
outputoff:
  html_document
output:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

## Titles

???

@@

Version 1: In statistics, comparing the effect of a treatment in two different
conditions requires the use of an appropriate statistical test to assert
the significance of interaction between the treatment and the condition.
However, a common procedure is to define genes specifically regulated in
one, but not the other condition by selecting genes which show a
statistically significant difference in one condition only. Here we show
that not only is this procedure incorrect, but moreover, when combined with
gene set enrichment analyses, it leads to statistically significant, but
incorrect results. These gene set enrichments are the effect of false
negatives and reflect only the differences in gene expression due to
treatment and not due to the difference between conditions. Thus, the
combination of incorrect statistical analysis and gene set enrichment
analysis may result in particularly misleading artifacts.

Version 2: A common application of differential expression analysis is to
ask which genes are differentially expressed upon treatment in one
condition only. One of the approaches is to define genes specific to a
condition by testing for significant difference in expression separately
in the two conditions, and then selecting genes which show statistical
significance in one condition only. This approach may then be combined with
a gene set enrichment analysis to find pathways and gene sets
characteristic for this condition only.

Here we show that this procedure is incorrect. Instead, the interaction
between treatment and condition should be tested. Moreover, we show that
gene set enrichment analysis applied to incorrectly defined genes specific
to one condition only may result in misleading artifacts. Genes which show
significant difference upon treatment in one, but not the other condition
are enriched in gene sets related to the treatment, but not to the
condition. Thus, while the results are apparently related to the treatment,
they do not reflect differences in expression specific to a given
condition.

Version 3:

@@

```{r setup}
padj_thr <- 0.05
lfc_thr  <- 1
n_stratum <- 10 # 2 strata (male and female) per condition/group
                # total number of samples n_stratum * 2 * 2 * 2
lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
```

## Introduction

Transcriptomic analyses commonly include more than one factor. Often, the
question asked is whether there is a difference between conditions (first
factor) with respect to reaction to a particular treatment (second
condition). For example, we may ask whether there are differentially
expressed genes (DEGs) which are specific to a particular group of
patients, such as interferon response elicited by a virus in one group, but
silenced in the other group of patients. In other words, we ask whether the
difference between control (healthy subjects) and treatment (infected
patients) is different between two groups of individuals. This "difference
of differences" is known in statistics as an interaction
[@blalock1965theory]. To find out whether it is statistically significant,
an appriopriate statistical test for interaction should be employed.

However, another approach is widely spread [@nieuwenhuis2011erroneous].
Instead of testing the interaction, one tests the effect of the treatment
separately in both groups. Next, a difference between conditions is inferred if the difference
is significant in one comparison, but not significant in the other. This
approach is not correct from statistical point of view, as "the difference
between significant and not significant is not itself statistically
significant" [@gelman2006difference]. For example, the p-value in the first
comparison may be 0.009, and in the other comparison 0.011, thus at an
alpha level of 0.01 the difference will be statistically significant in the
first, but not significant in the other comparison.

In transcriptomics, the statistical tests are performed for thousands of
genes. As in the general case, the correct inference of differences between
the conditions is by testing the significance of interaction. However, in
another approach the differences between treatment and control are tested
in the two conditions separately, and the results are then compared. This
can be visualized using a Venn diagram (VD, Fig. 1) showing the overall
number of DEGs significant in both comparisons (the intersection in the VD)
or significant in only one comparison (the remaining two fields on a VD).
The genes which are significant in only one comparison may then be
incorrectly defined as specific to the corresponding condition. Following
this, gene set enrichment analysis may be applied to these genes in an
attempt to test which pathways are specific to one, but not the other
condition.

In this paper, we show that under reasonable assumptions this approach may
result in apparent enrichments even if there are no real differences
between the groups. To this end, we have randomly split a cohort into two
groups, compared the treatment (viral infection) with controls in each of
the groups separately and then applied gene set enrichment to the sets of
genes significantly different in one, but not the other group.
Moreover, we show that the resulting gene set enrichments correspond to the
differential expression between treatment and control. Thus, the enriched
terms are relevant to the studied project, yet while they do reflect real
processes linked to treatment, they are not demonstrative of differences
between the study groups. Finally, we use literature search to show that
this incorrect approach to study group-specific treatment effects is not
uncommon.

@@

First, we have performed differential gene expression analysis for each of
the groups G1 and G2 separately using standard bioinformatic tools. For each comparison, we defined
DEGs
by using an false discovery rate (FDR) threshold of 
`r padj_thr`
and absolute log~2~ fold change (LFC) threshold of
`r lfc_thr`. There were
`r sum(res$g1$DEG)` 
DE genes in the G1 group, and
`r sum(res$g2$DEG)`
in the G2 group. In total,
`r sum(res.merged$DEG.g1 & res.merged$DEG.g2)` 
DEGs were common for G1 and G2, 
`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
DEGs were significant in G1 only ("specific" for G1), and 
`r sum(!res.merged$DEG.g1 & res.merged$DEG.g2)` 
were significant in G2 only (see Fig. 1, panel *A*).
A naive interpretation of these results implies that there is a substantial
difference between these two groups of individuals, as evidenced by a small
overlap in commonly regulated genes, while the majority of DEGs is
significant in one comparison only.

@@

```{r GSEA,cache=FALSE}
common <- res.merged %>% filter(DEG.g1 & DEG.g2) %>% pull(SYMBOL)
gsea_res <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset)
})
gsea_res_full <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset, qval = Inf)
})
message("G1 ", nrow(gsea_res$g1), " G2 ", nrow(gsea_res$g2))
gsea_res.merged <- merge(gsea_res$g1, gsea_res$g2, by=c("ID", "Title"), suffixes=c(".g1", ".g2"), all=T)
saveRDS(gsea_res.merged, file="gsea_res_merged.rds")
```

To understand which
pathways are upregulated in each of the two groups, we used a standard generation I
gene set enrichment analysis – a hypergeometric test – on the DEGs in each
group. Gene sets for the gene set enrichment analysis were taken from the Gene Ontology (GO) database.
Gene sets with more than `r max_mod_size` or 
fewer than `r min_mod_size` genes were removed.
For each group, we have
selected only genes which are DEGs in that group, but not the other,
mimicking a naive approach for finding pathways regulated in one patient
group only.
Here, a similar picture
emerged. Just
`r nrow(gsea_res$g1)` gene sets were significantly enriched in G1, and
`r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. Again,
both the Venn diagram (Fig. 1, B) and the results of
enrichments (Fig. 1, C and D) suggest that there is a
fundamental difference between the groups, and that the groups have little
in common in their response to the virus.

@@

These FNs will occur
at a much higher rate within the sets of DEGs defined by the
non-overlaping areas of the VDs, that is DEGs considered to be "specific"
for one group or other in a naive approach.
To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
were drawn (Fig. 3), comparing the 
`r sum(covar$disease == "no") `
healthy controls to 
`r sum(covar$disease == "SC2") `
COVID-19 patients. Of the
`r length(g1_spec)`
genes significant in G1, but not in G2, 
`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g1_spec) %>% nrow; .n`
(`r sprintf("%.0f%%", .n/length(g1_spec) * 100)`) 
are significant in the full data set; 
of the
`r length(g2_spec)`
genes significant in G2, but not in G1, 
`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g2_spec) %>% nrow; .n`
(`r sprintf("%.0f%%", .n/length(g2_spec) * 100)`) 
are significant in the full data set.
Given that G1 and G2 were sampled from the total population, and since the
FDR was set to `r padj_thr`, we can assume that at least between a third and a half of
the genes that appeared to be "specific" in the initial analysis were, in
fact, false negatives in one of the comparisons.

**Fig. 3. Area-proportional Venn diagram showing overlaps in DEGs between
G1, G2 and the full data set.** The majority of genes which have been labeled
as DEGs in only one of the groups G1, G2 are DEGs when all data were
analysed.

@@

In the example above, the groups have been randomly sampled from a larger
data set only once. Arguably, the observations might differ if the groups
were to be resampled. Furthermore, we have chosen a group size of 
`r n_stratum * 4` (`r n_stratum * 2` per group/treatment combination). 
Larger sample sizes are known to increase robustness of gene set enrichment
analysis, and group size of 20 has been shown to be relatively robust [@maleki2019size].
However, a smaller or larger group size might change the proportion of FNs in the
results and thus influence the results of gene set enrichment. Finally,
we have used a log~2~ fold change threshold of `r lfc_thr`, because raising
it would increase the fraction of FNs. However, selecting more biologically
only relevant genes which a substantially higher effect size may influence
the observed enrichments. On the other hand, raising the log~2~ fold change
threshold or lowering the sample size may lead to a smaller number of
defined DEGs, thus making the hypergeometric test less powerful and lead to
fewer gene set enrichment.

To investigate whether the sampling had an impact on
the artifacts, we have repeated the above procedure for 100 replicates,
each containing a different set of samples randomly assigned to the two
groups.  Furthermore, we tested whether the selection of the log~2~ fold
change threshold or different sample size might have an impact on the
extent of the arising artifacts. To this end, we have tested 
`r length(lfc_thresholds)` 
different threshold values for the log~2~ fold
changes and three sample sizes: 20, 40 or 80 samples per group
(corresponding to 10, 20 or 40 samples per group/treatment combination).

**Fig. 4. Influence of log~2~ fold change treshold (A, C) and the sample
size (B, D) on the number of genes significant in one, but not in the other
group (A, B) and the number of significant gene sets found in one group, but
not the other (C, D).**  Top row shows the influence of log~2~ fold change
threshold for sample size of 40 per group (20 per group/treatment
combination), bottom row shows the influence of sample size for a log~2~
fold change threshold set to `r lfc_thr`.

```{r montecarlo,cache=TRUE,fig.width=7.5,fig.height=6.5}
## note: given that the monte carlo simulation is quite time consuming, we
## have included the code for it in another file, `montecarlo.R`. If trying
## to replicate this manuscript, please execute that file first to generate
## the necessary replicates.

@@

fdr_thresholds <- c(.8, .7, .6, .5, .4, .3, .2, .1, .05, .01, .001, .0001, 0.00001, 1e-6)
by_fdr <- map_dfc(fdr_thresholds, ~ {
                      a <- get_specific_num(merged_res[["10"]], lfc_thr=0, padj_thr=.x)
                      b <- get_total_num(merged_res[["10"]], lfc_thr=0, padj_thr=.x)
                      (b - a)/b }
)
colnames(by_fdr) <- fdr_thresholds
by_fdr <- by_fdr %>% pivot_longer(everything(), names_to = "lfc")
by_fdr$lfc <- factor(by_fdr$lfc, levels=fdr_thresholds)
g0 <- ggplot(by_fdr, aes(x=lfc, y=value)) + geom_violin() +
  ylab("Number of DEGs") + xlab("FDR threshold")

@@

Setting a higher log~2~ fold change threshold reduces the number of
DEGs as well as of the observed artifactual gene set enrichments 
(Fig. 4, A and B). However, even for log~2~
threshold of 3 (DEGs defined by 8-fold change and FDR < `r padj_thr`) the
number of replicates in which artifacts can be observed is 
`r sum(gsea_by_lfc_thr %>% filter(lfc == 3) %>% pull(value) > 0)`
(out of 100),
and in at least 
`r sum(gsea_by_lfc_thr %>% filter(lfc == 3) %>% pull(value) >= 5)`
replicates, 5 or more gene sets were significantly
enriched. Thus, setting a more conservative threshold while retaining the
incorrect procedure cannot fully protect from the arising artifacts.
The number of artifacts rose with sample size
(Fig. 4, C and D). For total sample size of 160 (40 samples per
group/treatment combination) the number of artifacts was almost 
`r round(
  mean(gsea_by_ss %>% filter(ss == 40) %>% pull(value)) /
  mean(gsea_by_ss %>% filter(ss == 20) %>% pull(value)))
`
times higher than for total sample size of 80.

### Incorrect analysis of interactions is common in transcriptomics

@nieuwenhuis2011erroneous observed incorrect analyses of interactions in
about half of the analysed papers from top neuroscience journals where the
authors considered an interaction. We wanted to know if this problem is
common in transcriptomics, too. To this end, we have searched three
journals – from broad to specialized – for the occurence of the terms
"differential expression" with "venn diagrams" (Tab. 2). Next, we analysed
the selected papers from year 2020 (and years 2015-2020 in one case) to
decide whether the VD was described or referred to as showing genes
"specific" or "unique" to a particular condition or whether a test for
interaction was performed. Finally, we checked whether gene set enrichment
analysis was applied to genes significant in one, but not the other
condition in order to find gene sets and pathways regulated by one, but not
the other condition.

@@

We found 
that of the 
`r tab2$Analysed[nrow(tab2)]` analysed articles
 which used the terms "venn diagram" and
"differential expression", at least 
`r tab2$Incorrect[nrow(tab2)]`
were using Venn diagrams to compare
statistical significance with lack thereof by referring to "unique",
"specific", "solely regulated" or "exclusive" DEGs. Out of these, at least 
`r tab$Enrichment[nrow(tab2)]`
coupled the VDs with some form of gene set enrichment analysis on the
set of supposedly "specific" DEGs. In summary, in at least a quarter of the
papers on differential expression in which a venn diagram was used, it was
illustrating an incorrect statistical procedure which may result in
artifactual gene set enrichments.

Drawing conclusions from comparing significance with lack thereof is a
common statistical fallacy [@gelman2006difference]. Just as absence of evidence is not evidence of
absence, the failure to reject the null hypothesis does not consitute the
same level of evidence as rejecting it. However, when such an incorrect
analysis is combined with downstream functional analysis, the
resulting pathways or gene ontologies are misleadingly relevant. For
example, the identified gene sets are associated with immune answer for
research hypotheses involving an infectious disease, or cancer pathways if
the underlying research hypothesis involved cancer treatment. Such results
may appear reasonable in the given context, especially if the correct
analysis of interactions does not show any significant differences. This
effect is persistent or even exacerbated for larger sample sizes (Fig.
4).

We found that this type of incorrect analysis occurs in more than a quarter
of papers where the procedure was illustrated with a VD.  That is not to
say that VDs are not a useful tool, even in the context of transcriptomics
and gene set enrichments, if used correctly. For example, analysis of an
intersection of DEGs (i.e., by considering genes from the overlap in a VD)
is not an incorrect procedure.  Genes in the overlapping part of a VD are
significant in both (or all) comparisons, hence no comparison between
significance and non-significance is made.

As an alternative to Venn diagrams and the downstream gene set enrichment
analysis, two approaches can be considered.
The correct statistical approach, as shown above, involves a test for interaction which
can reveal genes for which the impact of treatment significantly differ
between the groups. The results can then be plugged into a gene set
enrichment analysis the usual way.  Unfortunately, this has two major
drawbacks. Firstly, effect sizes (log~2~ fold changes) of the interaction
term are harder to interpret than log~2~ fold changes in a direct, group vs
group comparison.  The effect size in an interaction is negative if the
log~2~ fold change in the first comparison is larger than the log~2~ fold
change in the second comparison; this is, however, irrespective of whether
the differences in the individual comparisons are negative or positive,
which makes it harder to separate the differences in genes up-regulated in
one or both groups.
